DCMIP2016, Part 1: Models and Equation Sets

### 2016-12-28 DCMIP2016-Part1-Models.tex
- Further cleanup and additions - Moved dynamical core overviews to top

The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.  Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) throughout the atmosphere, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.  These factors include the choice of model grid, vertical coordinates, representation of topography, numerical method, physics/dynamics coupler, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP has hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention is paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.

The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.  The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.  Modeling groups have already begun to leverage this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.

The proposed workshop has advanced our knowledge of (a) the relative behaviors exhibited by atmospheric dynamical cores, (b) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (c) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.

\section{Notation}

\subsection{List of Symbols}

Table \ref{tab:symbols} lists the symbols used in this paper.

\begin{table}[h]
\caption{List of symbols used in this manuscript} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\Theta$ & Potential temperature \\
$\Theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

\subsection{List of Physical Constants}
A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.  Constants which are specific to each test case are similarly tabulated at the beginning of each section.

\begin{table}[h]
\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
%\ \\
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
\hline Constant & Description & Value \\
\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
$c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
%$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
\hline 
\end{tabular*}

\end{table}

\subsection{Great Circle Distance}

The great circle distance is used throughout the document and is given by
\begin{equation}
R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
\end{equation}

@@

Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.  In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.  This section describes many of the model grids that have been employed globally.

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude {\color{red}[Figure]}.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new}.

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

@@

Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.  For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.  However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.  This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.  Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e. for ocean basins or ice sheets) \citep{ringler2008multiresolution}.  CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.  However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.

@@

As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for global 
medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

@@

\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}

In this section we describe the fluid equations utilized by nonhydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}

@@

@@

\subsection{GEM} \label{sec:GEM_equations}
The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:

>>>>

The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.  Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) throughout the atmosphere, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.  These factors include the choice of model grid, choice of prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to

The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.  The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.  Modeling groups have leveraged this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.

The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.

This paper is the first in a series of papers documenting the results of this workshop.  This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.  Section \ref{sec:Notation} describes the unified notation that our project has adopted that will be relevant throughout all papers in this series.  The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; section \ref{sec:TemporalDiscretizations} describes temporal discretizations; and section \ref{sec:DynamicalCores} provides an overview of how these techniques are tied together in each modeling system.

\section{Notation} \label{sec:Notation}

This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.  Table \ref{tab:symbols} lists the symbols used in this paper.  A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.  Great circle distance is used throughout the project and is computed via
\begin{equation}
R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
\end{equation}

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
%\ \\
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
\hline Constant & Description & Value \\
\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
$c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
%$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
\hline 
\end{tabular*}
\end{table}

\section{Dynamical Cores} \label{sec:DynamicalCores}

This section provide a short description of unique features or design specifications from several modern dynamical cores.  A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.

\begin{table}[h]
\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & Finite Volume Cubed Model & Geophysical Fluid Dynamics Laboratory \\
FVM & Finite Volume Module of the Integrated Forecast System & European Center for Medium-Range Weather Forecasting \\
GEM & Global Environmental Multiscale model & Environment Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
\hline 
\end{tabular}
\end{center}
\end{table}

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Time-differencing is based on the third-order Adams-Bashforth scheme. Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.  
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature,\citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
\label{sec:FVM}

The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
are available by means of simple switches. The formulation incorporates 
a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
Conjugate Residual approach. 
The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
The horizontal spatial discretization is fully unstructured finite-volume using the 
median-dual approach. This is combined with a structured-grid finite-difference approach in the 
vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
In both the horizontal and the vertical discretization, all prognostic variables are 
co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.

No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
at every time step according to a given maximum advective CFL number (typically somewhat smaller 
than 1). The physics time step is identical to the dynamics time step. 
%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.  The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.  A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.  A scalar viscosity is employed for wind components and tracers 
via an application of (\ref{eq:scalar_viscosity_direct}).  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}

The High Order Method Modeling Environment (HOMME) is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following / pressure coordinate. The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$. The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

@@

Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.  In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.  The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.  The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.  Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.  To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new}.

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001shallow}.  An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

@@

Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.  For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.  However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.  This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.  Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e. for ocean basins or ice sheets) \citep{ringler2008multiresolution}.  Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.  CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.  However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.

@@

As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for global 
medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

@@

\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}

In this section we describe the fluid equations utilized by non-hydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

%{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}

@@

\subsection{Vorticity-Divergence Form}

The vorticity-divergence form of the dynamical equations in the height vertical coordinate predicts the (vertical component of) vorticity and (velocity) divergence given by
\begin{align}
\omega_z \equiv \vb{k} \cdot \nabla_s \times \vb{u}_h,
\end{align} and
\begin{align}
D \equiv \nabla_s \cdot \vb{u}_h,
\end{align} respectively, instead of the horizontal velocity.

The vorticity prediction equation is obtained by taking the curl ($\vb{k} \cdot \nabla_s \times$) of the horizontal momentum equation as
\begin{align}
\pdiff{\omega_z}{t} + \nabla_s \cdot (\omega_z^\ast \vb{u}_h) + \pdiff{}{z} (\omega_z^\ast w) - \omega_z \pdiff{w}{z} - \pdiff{}{z} (\vb{k} \times \vb{u}_h) \cdot \nabla_s w + J(c_p \theta_v, \pi) = 0.
\end{align} The Jacobian   is a compact form for the curl of the pressure gradient force, where is the specific heat of air at constant pressure.

@@
